Surface-driven registration method for the structure-informed segmentation of diffusion MR images\tnoteref{t1}

### 2016-03-10 Publications/2015-NeuroImage/04-methods.tex
latest editions prior submission

\begin{figure*}
% \includestandalone[width=\linewidth]{figure02-model}
\begin{minipage}{\textwidth}
\includegraphics[width=\linewidth]{figure02}\\%
\vskip-2.0em
\includegraphics[width=\linewidth]{legend}%
\end{minipage}
% \includegraphics[width=\linewidth]{figure04B}
% \includegraphics[width=\linewidth]{figure04C}
\caption{%
Evolution of the segmentation model defined by the homogeneous regions $\Omega_l$,
  for one real dataset.
Panel (A, left) shows the joint distribution of the \gls*{fa} and \gls*{adc} conditioned to
  the segmentation \omegaset{} defined by the surfaces $\gammaset_R$ extracted from
  the \gls*{t1} image.
The plot was generated for reference using undistorted diffusion data, and
  therefore, $\gammaset_R$ is aligned with the \gls*{fa} and the \gls*{adc}.
The problem arises when the diffusion data present deformation, and the
  contours $\gammaset_R$ do not fit within the data (A, center).
After registration with \regseg{}, the contours are mapped onto the
  diffusion data (A, right), and the joint density plot is closer to the reference
  situation.
In panel (B), the three plots in (A) are decomposed tissue-wise.
Using filled contours, the bivariate distribution of each tissue is highlighted in its designated color,
  and represented over the remaining tissues (in gray colors).
To help assessment, dashed contours in black-to-white colors represent
  the corresponding distribution in the reference plot.
The registration process optimizes the segmentation model of \regseg{},
  and thus, the distribution of each region after registration is located
  closer to that corresponding in the reference situation, the shape of the distribution
  is more similar to the reference, and their spread is also reduced.
The effects of optimization are more noticeable on the \gls*{gm} ($\Omega_\text{GM}$) and
  the \gls*{wm} ($\Omega_\text{WM}$).
Particularly, the \gls*{wm} typically shows a bimodal distribution when the contours
  \gammaset{} do not fit the data.
% The left plot in each panel represents a kernel density estimate of the location and spread of
%   each region $\Omega_l$ in the bivariate feature space of $M$, which comprises
%   the \gls*{fa} and \gls*{adc} maps.
% At the top and the right margins, the marginal distributions of $\Omega_l$ on the \gls*{fa} and
%   \gls*{adc} are respectively displayed.
% On the right side, the sampled distribution of each region is shown in small scatter plot, in contrast to
%   the complete bivariate data.
% A) shows the initialization of the model, when the anatomically-correct surfaces are not
%   mapped into the \gls*{dmri} space.
% The panel shows the resulting conditional distributions of segmenting the (distorted) \gls*{fa} and
%   \gls*{adc} maps with the original (undistorted) surfaces.
% Panel B shows the equivalent information, using the surfaces that are perfectly mapped into the
%   distorted \gls*{dmri} space.
% Since the surfaces are now aligned in $M$, the spreads of the conditional distributions for each region are
%   reduced with respect to those in A.
% Panel C, after \regseg{} registration, shows how the covariance matrices $\boldsymbol{\Sigma}_l$
%   are minimized with respect to panel A, and the resulting segmentation is close
%   to the theoretical situation presented in panel B.}
}\label{fig:regseg-model}
\end{figure*}

@@

\paragraph*{Segmentation model}
Based on our experience % \citep{esteban_brain_2012}
  and previous studies \citep{ennis_orthogonal_2006},
  we defined the moving image as a stack of the \gls*{fa} and \gls*{adc} maps derived
  from \gls*{dmri} data.
After evaluating several alternative models, we empirically defined a partition \omegaset{}
  according to the following six regions:
  1) thalamus ($\Omega_{Tha}$);
  2) ventricular system and deep \gls*{gm} structures ($\Omega_{VdGM}$);
  3) cerebral \gls*{wm} ($\Omega_{WM}$);
  4) brain stem and cerebellar \gls*{wm} ($\Omega_{bst}$);
  5) cerebellar \gls*{gm} ($\Omega_{cbGM}$); and
  6) cortical \gls*{gm} ($\Omega_{GM}$).
Using tools in \emph{FreeSurfer} and appropriate selections of labels in the
  \emph{aparc} segmentation released with the \gls*{hcp} data, we extracted the $\gammaset_R$ set for the
  reference surfaces.
The segmentation model corresponding to this partition is shown in \autoref{fig:regseg-model}
  and discussed in greater detail in the \citetalias[section S4]{esteban_useful_2015}.

>>>>

\begin{figure*}
% \includestandalone[width=\linewidth]{figure02-model}
\begin{minipage}{\textwidth}
\includestandalone[width=\linewidth]{figure02}%
\end{minipage}
% \includegraphics[width=\linewidth]{figure04B}
% \includegraphics[width=\linewidth]{figure04C}
\caption{%
Evolution of the segmentation model defined by the homogeneous regions $\Omega_l$,
  for one real dataset.
Panel (A, left) shows the joint distribution of the \gls*{fa} and \gls*{adc} conditioned to
  the segmentation \omegaset{} defined by the surfaces $\gammaset_R$ extracted from
  the \gls*{t1} image.
The plot was generated for reference using undistorted diffusion data, and
  therefore, $\gammaset_R$ is aligned with the \gls*{fa} and the \gls*{adc}.
The problem arises when the diffusion data present deformation, and the
  contours $\gammaset_R$ do not fit within the data (A, center).
After registration with \regseg{}, the contours are mapped onto the
  diffusion data (A, right), and the joint density plot is closer to the reference
  situation.
In panel (B), the three plots in (A) are decomposed tissue-wise.
Using filled contours, the bivariate distribution of each tissue is highlighted in its designated color,
  and represented over the remaining tissues (in gray colors).
To help assessment, dashed contours in black-to-white colors represent
  the corresponding distribution in the reference plot.
The registration process optimizes the segmentation model of \regseg{},
  and thus, the distribution of each region after registration is located
  closer to that corresponding in the reference situation, the shape of the distribution
  is more similar to the reference, and their spread is also reduced.
The effects of optimization are more noticeable on the \gls*{gm} ($\Omega_\text{GM}$) and
  the \gls*{wm} ($\Omega_\text{WM}$).
Particularly, the \gls*{wm} typically shows a bimodal distribution when the contours
  \gammaset{} do not fit the data.
% The left plot in each panel represents a kernel density estimate of the location and spread of
%   each region $\Omega_l$ in the bivariate feature space of $M$, which comprises
%   the \gls*{fa} and \gls*{adc} maps.
% At the top and the right margins, the marginal distributions of $\Omega_l$ on the \gls*{fa} and
%   \gls*{adc} are respectively displayed.
% On the right side, the sampled distribution of each region is shown in small scatter plot, in contrast to
%   the complete bivariate data.
% A) shows the initialization of the model, when the anatomically-correct surfaces are not
%   mapped into the \gls*{dmri} space.
% The panel shows the resulting conditional distributions of segmenting the (distorted) \gls*{fa} and
%   \gls*{adc} maps with the original (undistorted) surfaces.
% Panel B shows the equivalent information, using the surfaces that are perfectly mapped into the
%   distorted \gls*{dmri} space.
% Since the surfaces are now aligned in $M$, the spreads of the conditional distributions for each region are
%   reduced with respect to those in A.
% Panel C, after \regseg{} registration, shows how the covariance matrices $\boldsymbol{\Sigma}_l$
%   are minimized with respect to panel A, and the resulting segmentation is close
%   to the theoretical situation presented in panel B.}
}\label{fig:regseg-model}
\end{figure*}

@@

\paragraph*{Segmentation model}
Based on our experience % \citep{esteban_brain_2012}
  and previous studies \citep{ennis_orthogonal_2006},
  we defined the moving image as a stack of the \gls*{fa} and \gls*{adc} maps derived
  from \gls*{dmri} data.
After evaluating several alternative models, we empirically defined a partition \omegaset{}
  according to the following six regions:
  1) thalamus ($\Omega_\text{Tha}$);
  2) ventricular system and deep \gls*{gm} structures ($\Omega_\text{VdGM}$);
  3) cerebral \gls*{wm} ($\Omega_\text{WM}$);
  4) brain stem and cerebellar \gls*{wm} ($\Omega_\text{bst}$);
  5) cerebellar \gls*{gm} ($\Omega_\text{cbGM}$); and
  6) cortical \gls*{gm} ($\Omega_\text{GM}$).
Using tools in \emph{FreeSurfer} and appropriate selections of labels in the
  \emph{aparc} segmentation released with the \gls*{hcp} data, we extracted the $\gammaset_R$ set for the
  reference surfaces.
The segmentation model corresponding to this partition is shown in \autoref{fig:regseg-model}
  and discussed in greater detail in the \citetalias[section S4]{esteban_useful_2015}.
